---
title: "Extending SigBridgeR: A Guide for Extend SCPreProcess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending SigBridgeR: A Guide for Extend SCPreProcess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SigBridgeR)
```

> This vignette will be deprecated as `SCPreProcess` is deprecated.

# Extend Seurat Methods

In general, any Seurat function can be registered for use within `SCPreProcess`. However, due to the character-length limitation of the pipeline code string, at most **52** functions can be registered—one for each uppercase and lowercase letter (A–Z, a–z).  

By default, the registry is defined as follows:

**Pipeline Code Table:**

| Code | Function               | Description                                      |
|------|------------------------|--------------------------------------------------|
| **o** | `CreateSeuratObject`   | **Required.** Must be the first step.            |
| **n** | `NormalizeData`        | Standard normalization.                          |
| **s** | `ScaleData`            | Scales data for PCA.                             |
| **v** | `FindVariableFeatures` | Selects highly variable genes.                   |
| **p** | `RunPCA`               | Principal Component Analysis.                    |
| **e** | `FindNeighbors`        | Computes SNN graph.                              |
| **c** | `FindClusters`         | Louvain algorithm clustering.                    |
| **t** | `RunTSNE`              | t-SNE reduction.                                 |
| **u** | `RunUMAP`              | UMAP reduction.                                  |
| **r** | `SCTransform`          | **SCT workflow.** Replaces n, s, v.              |

For example, if we need to use `Seurat::LoadXenium` to load Xenium data for screening, register this function first.

```{r}
#| label: extend_seurat
Register(x = Seurat::LoadXenium)
```

To check:

```{r}
#| label: check_seurat_extended
names(SCPreProcessStrategy)
```

```{r}
#| label: check_seurat_extended2
InterceptStrategy("SCPreProcessStrategy")
```

overwrite:

```{r}
#| label: overwrite_seurat
Register(x = Seurat::LoadXenium, overwrite = TRUE)
```

## Session Info

```{r}
#| label: sessionInfo
#| echo: false
sessionInfo()
```
