---
title: "Extending SigBridgeR: A Guide for Extend SCAnnotate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending SigBridgeR: A Guide for Extend SCAnnotate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r }
#| label: setup
library(SigBridgeR)
```

# Extend Cell Type Annotation Methods

By default, SigBridgeR includes three built-in cell type annotation algorithms: **SingleR**, **CellTypist**, and **mLLMCelltype**-all of which require additional installation and dependencies (See [README](https://wanglabcsu.github.io/SigBridgeR/index.html)). They can all be invoked through the unified interface function `SCAnnotate`.

Extending cell annotation method is easy:

```{r}
#| label: extend_cell_anno
my_method <- function(seurat_obj, ...) {
  # TODO: Annotate cell type
  return(seurat_obj)
}
my_sc_anno <- AnnotationMethod(
  method_name = "my_method",
  executor = my_method
)
Register(
  my_method = my_sc_anno
)
```

To check:

```{r}
#| label: extend_cell_anoo2
names(SCAnnotateStrategy)
```

```{r}
#| label: extend_cell_anoo3
InterceptStrategy("SCAnnotateStrategy")
```

Here are some recommendations for user-defined cell annotation functions.

1. The first input and the first output should both be Seurat objects. 
2. If multiple objects need to be returned, it is recommended to use a `list` format. And the first element of the list should be the Seurat object.

## Session Info

```{r}
#| label: sessionInfo
#| echo: false
sessionInfo()
```
