---
title: "Extending SigBridgeR: A Guide for Custom Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending SigBridgeR: A Guide for Custom Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| include: false
library(zeallot)
library(SigBridgeR)
```

This vignette introduces a guide for extending SigBridgeR with custom algorithms.

## Prerequisites

It's recommended to install these packages, as they will be used for code checking, validation, and the utility functions we will use.

```{r}
#| label: install
#| eval: false
pak::pkg_install(c(
  "tictoc",
  "yonicd/tidycheckUsage",
  # "codetools",
  "knitr",
  "lintr"
))
```

## Writing a Custom Algorithm

### Get a template

After the v3.2.0 update, SigBridgeR supports registering custom algorithms for screening phenotype-associated cell method into the package.

use `TemplateScreenFunc` to obtain the template, and use the function to write the workflow. Here, we create an algorithm named `foo`.

```{r}
#| label: template
tmpdir <- tempdir()
TemplateScreenFunc(
  filename = file.path(tmpdir, "foo.R"),
  func_name = "Foo",
)
```

```{r}
#| label: a_custom_algorithm
# Template screen function for SigBridgeR
#
# This file is generated by SigBridgeR::TemplateScreenFunc().
# Modify this function and register it with Register().

Foo <- function(
  matched_bulk,
  sc_data,
  phenotype,
  label_type = NULL,
  phenotype_class = c("binary", "survival", "continuous"),
  ...
) {
  # Validate phenotype_class parameter
  phenotype_class <- rlang::arg_match(phenotype_class)

  # Extract additional arguments
  dots <- rlang::list2(...)
  verbose <- dots$verbose %||% TRUE

  # TODO: Implement your screening logic here
  # Placeholder for modified scRNA data (replace with actual processing)
  modified_sc_data <- sc_data

  # Return result in expected format
  list(
    scRNA_data = modified_sc_data
    # You can add more items to the list if needed
  )
}
```

Format requirements for custom extension functions:  

**Input arguments**

The input arguments `matched_bulk`, `sc_data`, `phenotype`, `label_type = NULL`, and `phenotype_class`
must not be changed. This is to ensure the compatibility with the existing functions. Their roles are as follows:

- `matched_bulk` (required): takes a matrix. Gene symbols are used as row names, and samples are used as column names.
- `sc_data` (required): takes a `Seurat` object. 
- `phenotype` (required): 
  - A named vector if `phenotype_class` is `"binary"` or `"continuous"`. The `names` attribute matches the sample names (column names) of `matched_bulk`.
  - A data.frame if `phenotype_class` is `"survival"`. The column names must be `c("time", "status")`, and the row names should match the sample names (column names) of `matched_bulk`.
    - time: No specific format is required for the time variable; days, months, or years are all acceptable. 
    - status: Coded as 0 or 1 (0: censored, 1: event)
- `label_type` (required): A character used to label cell with study cases.
- `phenotype_class` (required): One or more of `"binary"`, `"survival"`, `"continuous"`.

**Output arguments**

- `scRNA_data` (required): A **Seurat** object with meta.data modified,
  - For columns added in the meta.data slot, it is recommended to **use the method name as a prefix** (e.g. foo) to distinguish them from those added by other methods.
- Other elements are optional. Necessary intermediate data can be returned.

The list returned by the custom function will be captured and used to create a `ScreenMethodResult` object.

------

If still have questions, use `LoadRefData` to load the reference data and check the format.

```{r}
#| label: load_ref_data
#| eval: false
#| 
ref %<-% LoadRefData(data_type = "binary")
seurat_bi <- ref[[1]]
bulk_bi <- ref[[2]]
pheno_bi <- ref[[3]]

ref2 %<-% LoadRefData(data_type = "survival")
seurat_sur <- ref2[[1]]
bulk_sur <- ref2[[2]]
pheno_sur <- ref2[[3]]
```

### Record algorithm parameters

After the update to v4.0.0, parameter recording is more standardized. You need to create all parameters as a `ScreenMethodConfig` object.

```{r}
#| label: a_ScreenMethodConfig
#| eval: false
config <- ScreenMethodConfig(
  method_name = "foo",
  param = list(
    foo_param_1 = 1L,
    foo_param_2 = "foo_param_2"
  ),
  method_version = "v0.1.0",
  phenotype_class = c("binary", "continuous", "survival"),
  label_type = "bar"
)
```

What has been recorded?

The name and version of your algorithm, the parameter list `param`, the phenotype classes supported by the algorithm, and the actual study label are recorded. These essentially make up the necessary information for a screening process and can be added to the `@misc` slot of the `Seurat` object.

```{r}
#| eval: false
#| label: add_to_misc
SeuratObject::Misc(modified_sc_data, "foo") <- config
```

or using `AddMisc`

```{r}
#| eval: false
#| label: add_to_misc_2
modified_sc_data <- AddMisc(modified_sc_data, foo = config)
```

### If you are developing a package

If you want to do more than simply write an algorithm function and instead organize your code as an R package, 
we also welcome-and even recommend-this approach. But please add the following code to the template function:

```{r}
#| eval: false
#| label: if_used_in_pkg
# If submitted to CRAN or bioconductor,
# you can use the following code to check if the package is installed
rlang::check_installed("<your-pkg>")
```

```{r}
#| eval: false
#| label: if_used_in_pkg_github
# If you are still developing
# you can use the following code to check if the package is installed from github
rlang::check_installed("pak")
rlang::check_installed("<your-repo>", action = \(pkg, ...) {
  pak::pak("<your-name>/<your-repo>")
})
```

The file containing this function can be placed under `inst/SigBridgeR` or under `R/`. If it is placed under `R/`, you need to follow the function export rules and add documentation. You can use the argument `documentation = TRUE` to generate some documentation automatically.

```{r}
#| label: if_used_in_pkg_documentation
TemplateScreenFunc(
  filename = file.path(tmpdir, "foo2.R"),
  func_name = "Foo",
  documentation = TRUE
)
```

```{r}
#| label: if_used_in_pkg_documentation_code
#| include: false
cli::cli_code(readLines(file.path(tmpdir, "foo2.R")))
```

### Validate the function

To facilitate format validation, we provide a function `ValidateScreenFunc` to check the above requirements. The validation output resembles that of `rcmdcheck`; please ensure there are no errors or warnings, and as few notes as possible.

```{r}
#| label: validation
ValidateScreenFunc(Foo)
```

As you can see, some variables in the template function are not used and are therefore detected.

By the way if providing a bad function

```{r}
#| label: bad_function
bad_fun <- function(x, save_path = "./analysis") {
  z <- x + 1
  y
  dir.create(save_path)
  return(NULL)
}

ValidateScreenFunc(bad_fun)
```

## Registering the function

### Temporarily add it to the package

Now we can register the function to the package: create a `ScreenMethod` object, and then register it to the package.

```{r}
#| label: register
my_foo <- ScreenMethod(
  method_name = "Foo",
  method_version = "0.1.0",
  executor = Foo,
  phenotype_class = c("binary", "continuous", "survival"),
  mapper = NULL
)

Register(Foo = my_foo)
```

Here, `mapper` can accept a function. This function must take exactly one argument, whose type must be `list`, and it must also return a `list`. The `mapper` can be used to transform certain parameters.

`Register` must be passed a **key-value pair**: the key serves as the registration name, and the value is the registered entity.

Let's check whether it has indeed been registered 

```{r}
#| label: check_registry_2
names(ScreenStrategy)
```

```{r}
InterceptStrategy("ScreenStrategy")
```

### Long-term support in SigBridgeR

After completing the steps described before, what you need to do is

1. Create a R file in `R/`, and name it in format `21{k}-Do{method_name}.R`, where `{k}` is letter `a` to `z`, and `{method_name}` is the name of your method.
  - Declare your contribution, modification time, the current SigBridgeR version, and a brief introduction. You may also include any other reasonable information, like `# nocov start` head.
  - Organize this information as a YAML header and place it at the beginning of the file.
  
2. Modify [70-ScreenStrategy.R](https://github.com/WangLabCSU/SigBridgeR/blob/main/R/70-ScreenStrategy.R), and fill in the function name you just wrote into `ScreenStrategy`
   
3. Add your contribution information to `README.Rmd`.

4. Perform project-level code formatting, documentation, and `R CMD CHECK`.
 
5. Create a pull request.

See [CONTRIBUTING.md](https://github.com/WangLabCSU/SigBridgeR/blob/main/CONTRIBUTING.md) to learn more about the further workflow to contribute to SigBridgeR.

## Use the function

Now we can use the function in the `Screen` function:

```{r}
#| label: use_function
#| eval: false
a_seurat <- SeuratObject::CreateSeuratObject(Matrix::Matrix(
  1:100,
  nrow = 10,
  dimnames = list(paste0("gene", 1:10), paste0("cell", 1:10))
))

bulk <- matrix(
  1:100,
  nrow = 10,
  dimnames = list(paste0("gene", 1:10), paste0("sample", 1:10))
)

pheno <- setNames(sample(0:1, 10, TRUE), paste0("sample", 1:10))

my_res <- Screen(
  sc_data = a_seurat,
  matched_bulk = bulk,
  phenotype = pheno,
  label_type = "Test",
  phenotype_class = "binary",
  screen_method = "Foo",
  a = 123
)
```

If you still have questions, please use GitHub [Issues](https://github.com/WangLabCSU/SigBridgeR/issues) or [Discussions](https://github.com/WangLabCSU/SigBridgeR/discussions).

## Other development guideline

In addition to the code formatting and interface requirements described above, there are several tips for developing a new algorithm:

- **CODE STYLE**: [tidyverse style](https://style.tidyverse.org/index.html) is recommended.

- **PARAMETER**: Instead of passing raw arguments directly into your algorithm, **construct a dedicated parameter object** and perform validation and preprocessing inside that object's builder. Doing the checks and normalization up front keeps your algorithm clean and prevents malformed inputs from reaching the screening logic. Only after the parameter object has been validated should you proceed to run the algorithm.

- **PERFORMANCE**: If your algorithm is computationally expensive, **prioritize optimizing the code itself before reaching for parallelization**. Parallelization can multiply resource usage and complexity, and it often masks inefficiencies that a better implementation would remove. Consider the following strategies, roughly in order: 
    1. Use high-performance R packages and functions;
       - Packages:
         - [cheapr](https://github.com/NicChr/cheapr)
         - [data.table](https://github.com/Rdatatable/data.table)
         - [matrixStats](https://github.com/HenrikBengtsson/matrixStats)
         - [sparseMatrixStats](https://github.com/const-ae/sparseMatrixStats)
         - [qs2](https://github.com/qsbase/qs2)
         - [matrixTests](https://github.com/karoliskoncevicius/matrixTests)
       - Functions: 
         - `stats::cor` -> `[WGCNA](https://github.com/cran/WGCNA)::cor`
    2. Rewrite the hot path in a high-performance language such as C++, Rust, or Julia;
    3. Consider hardware accelerators such as GPU or TPU.

- **FILE STORE**: If non-temporary, meaningful intermediate or analysis files need to be written, organize them in a `{method_name}_res` directory.

- **RESULT REPORT**: The result returned by your algorithm **should assign a specific label** to every cell. This can be delivered as a simple vector (if you favor a lightweight, minimal development) or as a data.frame (more comprehensive and easier to extend and reuse). Avoid returning only a phenotype-association score, because a bare score is difficult to apply directly in concrete studies. Such a score is not suitable as research material on its own; on the contrary, it requires the results of a study to corroborate it.

## Session Info

```{r}
#| label: sessionInfo
#| echo: false
sessionInfo()
```
