---
title: "Extending SigBridgeR: A Guide for Custom Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending SigBridgeR: A Guide for Custom Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document introduces a guide for extending SigBridgeR with custom algorithms.

# Too Long; Didn't Read

Extend `SCPreProcess`: key = func

```{r,tl_dr,eval= FALSE}
Register(h = Seurat::RunHarmony, "SCP")
```

Extend `SCAnnotate`: key = func

```{r,tl_dr2,eval= FALSE}
Register(my = my_function, "SCA")
```

Extend `Screen`: key = func

```{r,tl_dr3,eval= FALSE}
ValidateScreenFunc(my_function)
Register(my = my_function, "Scr")
```

# Extend Screening Methods

## Installation

It's recommended to install these packages for checking the code

```{r, install,eval= FALSE}
pak::pkg_install(c(
  "tictoc",
  "yonicd/tidycheckUsage",
  "codetools",
  "knitr",
  "lintr"
))
```

## Prepare a custom function

```{r, include = FALSE, eval= FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,setup, eval= FALSE}
library(SigBridgeR)
```

After the v3.2.0 update, SigBridgeR supports registering custom algorithms for screening phenotype-associated cell method into the package. Let's do this with a detailed example:

(Template can be obtained via `TemplateScreenFunc`, which will create a file and open it in the editor.)

```{r, a_custom_algorithm,eval= FALSE}
my_screen_function <- function(
  matched_bulk,
  sc_data,
  phenotype,
  label_type = NULL,
  phenotype_class = c("binary", "survival", "continuous"),
  ...
) {
  dots <- list(...)
  verbose <- dots$verbose %||% TRUE
  # do something, here we just randomly assign a label to each cell
  modified_sc_data <- SeuratObject::AddMetaData(
    sc_data,
    c(
      rep("Positive", floor(ncol(sc_data) / 2)),
      rep("Negative", ceiling(ncol(sc_data) / 2))
    ),
    col.name = "my_method"
  ) %>%
    # record parameters
    AddMisc(
      my_method_label = label_type,
      phenotype = phenotype_class
    )

  intermediate_var <- "value"

  if (verbose) {
    cli::cli_alert_success("my_screen_function finished")
  }

  list(
    scRNA_data = modified_sc_data,
    intermediate_var = intermediate_var
  )
}
```

Format requirements for custom extension functions:  

1. The input arguments must include 
   - `sc_data` (required): A **Seurat** object
   - `matched_bulk` (required): A **data.frame/matrix/Matrix** (**genes × samples**) containing RNA-seq counts:
     - Genes must overlap with those in `sc_data`.
     - Samples must correspond to those in `phenotype`.
   - `phenotype` (required):
     - For survival: a **data.frame** with columns `time` and `status`; rownames must match `colnames(matched_bulk)`
     - For binary or continuous: a **named vector**; names must match `colnames(matched_bulk)`
   - `label_type` (required): A character used to label cell with study cases.
   - `phenotype_class` (required): One or more of `"binary"`, `"survival"`, `"continuous"`.
  
2. The output must be a **list** containing at least one elements:
   - `scRNA_data` (required): A **Seurat** object with meta.data modified
   - Other elements are optional.

To facilitate format validation, we provide a function `ValidateScreenFunc` to check the above requirements. The validation output resembles that of `rcmdcheck`; please ensure there are no errors or warnings, and as few notes as possible.

```{r, validation,eval= FALSE}
ValidateScreenFunc(my_screen_function)
```

```{r, validation_2,eval= FALSE}
# ── Screening Function Validation ──────────────────────────────────────────────────────────────────────
# Start at 2026/01/19 22:04:17

# ✔ All input arguments explicitly specified

# ✔ Verbose control supported

# ✔ Syntax check passed

# ✔ Return value is a list with `scRNA_data` slot

# Duration: 0.095 sec elapsed

# 0 error ✔ | 0 warning ✔ | 0 note ✔

```

By the way if providing a bad function

```{r,bad_function,eval= FALSE}
bad_fun <- function(x) {
  z <- x + 1
  y
  return(NULL)
}

ValidateScreenFunc(bad_fun)
```

```{r,bad_function_2,eval= FALSE}
# ── Screening Function Validation ───────────────────────────────────────────────────────────────────
# Start at 2026/01/19 22:07:43

# ❯ Missing required arguments ... ERROR
#   More arguments should be added:
#   sc_data: A fully preprocessed Seurat object
#   matched_bulk: Bulk RNA-seq matrix (gene * samples):
#                 • Samples match `phenotype` in number/order;
#                 • Senes overlap with `sc_data`
#   phenotype: Phenotype: a named vector or data.frame, names/rownames match `matched_bulk`
#              • For binary/continuous: named vector recommended
#              • For survival: data.frame with `time` (1st col) and `status` (2nd col) recommended.
#   label_type: Labeling phenotype-associated cell with real study identifiers
#   phenotype_class: Phenotype types:
#                    • Must be one or more of `binary`, `continuous` and `survival`

# ❯ Verbose control not supported ... NOTE
#   Consider adding `verbose` control to ease error tracing

# ❯ Syntax error in function ... ERROR

# | line | object | col1 | col2 |   warning_type    |                     warning                     |
# |:----:|:------:|:----:|:----:|:-----------------:|:-----------------------------------------------:|
# |  2   |   z    |  3   |  3   |   unused_local    | local variable ‘z’ assigned but may not be used |
# |  3   |   y    |  3   |  3   | no_global_binding |   no visible binding for global variable ‘y’    |

# ❯ Return value is not a list ... ERROR
# scRNA_data: recommended to be the first element of the return value
#             • Should be of class <Seurat>

# Duration: 0.158 sec elapsed

# 3 errors ✖ | 0 warning ✔ | 1 note ✖

```

## Registering the function

Now we can register the function to the package:

```{r, register,eval= FALSE}
RegisterScreenMethod(
  my_method = my_screen_function,
  supported_phenotypes = c("binary", "survival"),
  parameter_mapper = function(params) {
    params$a <- params$a %||% 123
    params
  },
  registry = ScreenStrategy,
  verbose = TRUE
)
```

```{r,register_2,eval= FALSE}
# ✔ Registered `my_method`
```

Details of the arguments:

1. `my_method = me_screen_function`: 
   - formatting as **key = func**. Key is used to name the function. If no name provided, the function name will be used.
2. `supported_phenotypes`: The phenotype types supported by the function.
3. `parameter_mapper`: A function that transforms the input parameter list before passing it to the executor. Useful for changing parameters from interface function. **Receives a named list and must return a modified list**.
4. `registry`: The registry to register the function.
5. `verbose`: Whether to print messages.

Let’s check whether it has indeed been registered 

```{r, check_registry,eval= FALSE}
tbl <- InterceptStrategy("ScreenStrategy")
# A tibble: 22 × 4
```

```{r, check_registry_2,eval= FALSE}
names(ScreenStrategy)
# [1] "Scissor"   "scPP"      "LP_SGL"    "my_method" "PIPET"     "DEGAS"     "scAB"      "scPAS"
```

## Use the function

Now we can use the function in the `Screen` function:

```{r, use_function,eval= FALSE}
a_seurat <- SeuratObject::CreateSeuratObject(Matrix::Matrix(
  1:100,
  nrow = 10,
  dimnames = list(paste0("gene", 1:10), paste0("cell", 1:10))
))

bulk <- matrix(
  1:100,
  nrow = 10,
  dimnames = list(paste0("gene", 1:10), paste0("sample", 1:10))
)

pheno <- setNames(sample(0:1, 10, TRUE), paste0("sample", 1:10))

my_res <- Screen(
  sc_data = a_seurat,
  matched_bulk = bulk,
  phenotype = pheno,
  label_type = "Test",
  phenotype_class = "binary",
  screen_method = "my_method",
  a = 123
)
```

```{r, use_function_2,eval= FALSE}
# ✔ my_screen_function finished
```

```{r, use_function_3,eval= FALSE}
my_res$scRNA_data |> class()
```

```{r, use_function_4,eval= FALSE}
# [1] "Seurat"
# attr(,"package")
# [1] "SeuratObject"
```

If you still have questions, please use GitHub [Issues](https://github.com/WangLabCSU/SigBridgeR/issues) or [Discussions](https://github.com/WangLabCSU/SigBridgeR/discussions).

# Extend Seurat Methods

In general, any Seurat function can be registered for use within `SCPreProcess`. However, due to the character-length limitation of the pipeline code string, at most **52** functions can be registered—one for each uppercase and lowercase letter (A–Z, a–z).  

By default, the registry is defined as follows:

**Pipeline Code Table:**

| Code | Function               | Description                                      |
|------|------------------------|--------------------------------------------------|
| **o** | `CreateSeuratObject`   | **Required.** Must be the first step.            |
| **n** | `NormalizeData`        | Standard normalization.                          |
| **s** | `ScaleData`            | Scales data for PCA.                             |
| **v** | `FindVariableFeatures` | Selects highly variable genes.                   |
| **p** | `RunPCA`               | Principal Component Analysis.                    |
| **e** | `FindNeighbors`        | Computes SNN graph.                              |
| **c** | `FindClusters`         | Louvain algorithm clustering.                    |
| **t** | `RunTSNE`              | t-SNE reduction.                                 |
| **u** | `RunUMAP`              | UMAP reduction.                                  |
| **r** | `SCTransform`          | **SCT workflow.** Replaces n, s, v.              |

For example, if we need to use `Seurat::LoadXenium` to load Xenium data for screening, register this function first.

```{r,extend_seurat,eval= FALSE}
RegisterSeuratMethod(x = Seurat::LoadXenium)
# ✔ Registered x
```

To check:

```{r,check_seurat_extended,eval=FALSE}
names(SCPreProcessStrategy)
#  [1] "t" "u" "v" "x" "c" "e" "i" "n" "o" "p" "r" "s"
```

```{r,check_seurat_extended2,eval=FALSE}
tbl <- InterceptStrategy("SCPreProcessStrategy")
# tibble [12 × 2]
```

# Extend Cell Type Annotation Methods

By default, SigBridgeR includes three built-in cell type annotation algorithms: **SingleR**, **CellTypist**, and **mLLMCelltype**—all of which require additional installation and dependencies (See [README](https://wanglabcsu.github.io/SigBridgeR/index.html)). They can all be invoked through the unified interface function `SCAnnotate`.

Extending cell annotation method is also very easy:

```{r,extend_cell_anno,eval=FALSE}
my_method <- function(seurat_obj, ...) {
  # placeholder
  return(seurat_obj)
}

RegisterAnnoMethod(
  my = my_method
)
# ✔ Registered `my`
```

To check:

```{r,extend_cell_anoo2,eval=FALSE}
names(SCAnnotateStrategy)
# [1] "my"           "mLLMCelltype" "CellTypist"   "SingleR"
```

Here are some recommendations for user-defined cell annotation functions.

1. The first input and the first output must both be Seurat objects. 
2. If multiple objects need to be returned, it is recommended to use a `list` format.
3. Use `tidycheckUsage::tidycheckUsage()` to check the function's syntax.