---
title: "Auxiliary Utils"
date: '`r Sys.Date()`'
output: 
  html_document
knit: knitted
vignette: >
  %\VignetteIndexEntry{Auxiliary Utils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| eval: false
library(zeallot)
library(SigBridgeR)
```

This document introduces several auxiliary functions of SigBridgeR.

# Setting

## Setting and Retriving Package Options

Currently, 3 package options are provided:

- `verbose`: Whether to print the progress of the function. Default is `TRUE`.
- `seed`: The random seed used in the function. Default is `123L`.
- `timeout`: The maximum timeout time when downloading data. Default is `180L`.

These options can be modified using either the `setFuncOption` or `Options` function (the former is recommended).

```{r}
#| label: package_options_example
#| eval: false
setFuncOption(seed = 321L)

options(SigBridgeR.seed = 321L)
```

`setFuncOption` automatically detects the prefix, so parameters with a package name prefix are also compatible.

```{r}
#| label: package_options_example_2
#| eval: false
setFuncOption(SigBridgeR.seed = 321L)
```

`getFuncOption` is used to retrieve the global parameters of SigBridgeR.

```{r}
#| label: package_options_example_3
#| eval: false
getFuncOption("seed")
# 321

# * Auto-detect prefix
getFuncOption("SigBridgeR.seed")
# 321
```

## Setting Threads

`setThreads()` configures thread counts for OpenMP, data.table, and TensorFlow backends. By default, it uses **half of available physical cores** to balance performance and system responsiveness.

### Basic Usage

```{r}
#| eval: false
# Set explicit thread count
setThreads(8L)

# Configure specific backends only
setThreads(dt = 4L) # data.table only
setThreads(openmp = 4L) # OpenMP only
```

### TensorFlow Optimization
⚠️ **Must configure BEFORE importing TensorFlow**:

```{r}
#| eval: false
setThreads(
  threads = 8L,
  tf_config = list(
    inter_op = 2L,
    intra_op = 8L
  )
)
tf <- reticulate::import("tensorflow") # Import AFTER configuration
```

### Options
- `verbose = FALSE`: suppress console output
- Returns invisible list of old/new values for programmatic use
- Additional `data.table::setDTthreads()` arguments accepted via `...` (e.g., `restore = TRUE`)

> 💡 **Tip**: Default half-core allocation prevents oversubscription. Use full cores (`threads = availableCores()`) only on dedicated compute nodes.

# Cell Type Annotation 

As of version 3.3.0, three built-in cell annotation algorithms are included and can be applied directly to Seurat objects. Using them requires installing additional dependencies. See the [README](https://wanglabcsu.github.io/SigBridgeR/index.html).

Basic usage:

```{r}
#| label: cell_type_annotation_example
#| eval: false
seurat <- SCAnnotate(
  sc = seurat,
  method = c("CellTypist", "SingleR", "mLLMCelltype")
  # * more arguments passed to each method
)
```

`SCAnnotate` will automatically set up the required dependencies for you (making it even more plug-and-play). If the `method` argument is not specified, it will automatically select the appropriate annotation method based on the provided arguments.

To control it manually, use following functions:

## CellTypist Annotation

This is the direct R wrapper of celltypist.annotate. Unlike `SCAnnotate`, it does not prepare an environment for you.

```{r}
#| label: cell_typist_annotation_example_2
#| eval: false
seurat <- CellTypistAnnotate(
  sc,
  model = NULL,
  download = TRUE,
  conda = NULL,
  python = NULL,
  verbose = getFuncOption("verbose"), # TRUE
  celltypist_tools = system.file(
    "python/73-CellTypistAnnotate.py",
    package = "SigBridgeR"
  ),
  # * more arguments pass to celltypist.annotate (python)
)
```

**Parameters**:

- `sc`: A Seurat object.
- `model`: Charater/str. The model to use. If `NULL`, the default model will be used.
- `download`: Logical/bool. Whether to download the model. Defaults to `TRUE`-download all models. if model already exists, skip download.
- `force_update`: Logical. Whether to force update the model file.
- `conda`: Character/str. conda environment name. If `NULL`, try to use `python`.
- `python`: Character/str. The path to the python executable. If `NULL`, try to use `conda`.
- `verbose`: Logical/bool. Whether to print verbose messages. Defaults to `TRUE`.
- `celltypist_tools`: The path to the celltypist_tools.py file. If `NULL`, the default path will be used.

**Returns**:

A Seurat object with the cell type annotation added as metadata columns.

Other parameters supported by `celltypist.annotate` will be automatically passed through and can be directly supplied to `CellTypistAnnotate`.

helpful link: [https://github.com/Teichlab/celltypist](https://github.com/Teichlab/celltypist)

## mLLMCelltype Annotation

```{r}
#| label: mLLMCelltype_annotation_example
#| eval: false
seurat <- mLLMCellTypeAnnotate(
  sc,
  seurat_obj_markers = NULL,
  # * Context for large language models
  tissue_name = "Human Cancer",
  models = c(
    "gpt-5",
    "claude-sonnet-4-5-20250929",
    "gemini-3-pro",
    "qwen-max-2025-01-25"
  ),
  api_keys = list(
    anthropic = "your-anthropic-key",
    openai = "your-openai-key",
    gemini = "your-google-key",
    qwen = "your-qwen-key"
  ),
  # Other arguments passed to `mLLMCelltype::annotate_cell_types` or ` mLLMCelltype::interactive_consensus_annotation`
)
```

**Parameters**

- `sc` : A Seurat object with pre-computed clusters (stored in `Idents(sc)` or `sc$seurat_clusters`).
- `seurat_obj_markers` : Optional pre-computed marker gene table or list (output of `Seurat::FindAllMarkers()`). If `NULL` (default), markers are computed automatically using parameters passed via `...`.
- `tissue_name` : Character. Biological context for annotation (e.g., tissue type, disease state). Helps LLMs interpret marker genes appropriately. Default: `"Human Tumor"`.
- `models` : Character vector of LLM model identifiers. Supported formats:
    - OpenAI: `"gpt-4o"`, `"gpt-4o-mini"`, etc.
    - Anthropic: `"claude-3-5-sonnet-20240620"`, etc.
    - Google: `"gemini-1.5-pro"`, etc.
    - Alibaba: `"qwen-max"`, `"qwen-plus"`, etc.
    
    Default: `c("gpt-4o", "claude-3-5-sonnet-20240620", "gemini-1.5-pro", "qwen-max")`. For single-model mode, only the first model is used.
- `api_keys` : Named list of API keys with provider names as keys:
    - `openai`: OpenAI API key
    - `anthropic`: Anthropic API key
    - `gemini`: Google Cloud API key (with Gemini enabled)
    - `qwen`: Alibaba DashScope API key
    
    Example: `list(openai = "sk-...", anthropic = "sk-ant-...")`.
    
    **Note**: Default placeholder keys (`"your-xxx-key"`) will fail—users must supply valid keys.


**Returns**:

A Seurat object with the cell type annotation added as metadata columns.

helpful link: [https://github.com/cafferychen777/mLLMCelltype](https://github.com/cafferychen777/mLLMCelltype)

## SingleR Annotation

```{r}
#| label: singleR_annotation_example
#| eval: false
seurat <- SingleRAnnotate(
  sc,
  verbose = getFuncOption("verbose"), # TRUE
  # * pass to `celldex::HumanPrimaryCellAtlasData`
  ensembl = FALSE,
  cell.ont = c("all", "nonna", "none"),
  legacy = FALSE,
  # * pass to `SingleR::SingleR`
  ref = c("HPCA", "custom"),
  labels = NULL,
  # Other arguments passed to `SingleR::SingleR`
)
```

**Parameters**

- `sc`: A Seurat object.
- `verbose`: Logical. Whether to print progress messages. Defaults to `TRUE`.
- `ensembl`: Logical. Whether to convert row names to Ensembl IDs. Genes without a valid mapping are discarded.
- `cell.ont`: Character. specifying Cell Ontology term handling. Use `"nonna"` to discard samples without valid terms, `"all"` to keep all samples (with possible `NA`), or `"none"` to skip.
- `legacy`: Logical. Whether to use legacy ExperimentHub backend. Default uses gypsum.
- `ref`: Reference data for annotation. Either `"HPCA"` (Human Primary Cell Atlas from `celldex`) or a `SingleCellExperiment` object with pre-labeled cells. For multiple references, a list of matrices or `SummarizedExperiment` objects.
- `labels`: Character vector of cell type labels for `ref` columns. Required for custom references; defaults to `ref$label.main` for `"HPCA"`. If `ref` is a list, `labels` must be a corresponding list.

**Returns**:

A Seurat object with the cell type annotation added as metadata columns.

helpful link: [https://github.com/SingleR-inc/SingleR](https://github.com/SingleR-inc/SingleR)

# Setting up Python Environment

Some screening methods (e.g. [Section 3.5 DEGAS](#35-option-e-degas-screening)) are built using Python and require an execution environment. Here is a function to help you set up a Python environment. Both Windows and Unix-like systems are supported.

```{r}
#| label: setup_python_env_example
#| eval: false
# * This is an example of setting up a Python environment using conda
SetupPyEnv(
    env_type = "conda",
    env_name = "test-condaenv",
    method = c("reticulate", "system","environment"), # choose one of them, default is "reticulate"
    env_file = NULL # path to environment.yml file, used when method = "environment"
    python_version = NULL,
    packages = c(
        "pandas" = "1.3",
        "numpy" = "any"
    ),
    recreate = FALSE, # whether to remove the existing environment and recreate it
    use_conda_forge = TRUE,
    verbose = TRUE
) 

reticualte::use_condaenv("test-condaenv")
```

```{r}
#| label: setup_python_env_example_2
#| eval: false
# * Or use virtualenv via reticulate
SetupPyEnv(
  env_type = "venv",
  env_name = "test-venv",
  python_version = "3.9.15",
  packages = c("tensorflow" = "2.4.1", "protobuf" = "3.20.3"),
  python_path = NULL,
  recreate = FALSE,
  verbose = TRUE
)

reticulate::use_virtualenv("test-venv")
```

You can use `ListPyEnv()` to list all the Python environments you have set up. Both conda and virtual environments are supported.

```{r}
#| label: list_py_env_example
#| eval: false
# * Unix-like systems
ListPyEnv()
#                 name                                                  python  type
# 1               base                        /home/user/miniconda3/bin/python conda
# 2      test-condaenv     /home/user/miniconda3/envs/test-condaenv/bin/python conda
# 3          test-venv         /home/user/miniconda3/envs/test-venv/bin/python  venv
```

Show the conda environments only:

```{r}
#| label: list_py_env_example_2
#| eval: false
# * Unix-like systems
ListPyEnv(env_type = "conda")
#                 name                                                  python  type
# 1               base                        /home/user/miniconda3/bin/python conda
# 2      test-condaenv     /home/user/miniconda3/envs/test-condaenv/bin/python conda
```

If the virtual environment isn't installed in the default location, you can specify the location of the virtual environment with the `venv_locations` parameter.

```{r}
#| label: list_py_env_example_3
#| eval: false
ListPyEnv(env_type = "venv", venv_locations = "~/here_is_a_dir/.virtualenvs")
```

# Other Functions

## Load reference data

**Parameters**

- `data_type`: The type of data to load. Can be either "continuous", "survival" or "binary", case-insensitive.
- `path`: The path to the data directory. 
- `cache`: Whether to cache the data. Defaults to `TRUE`.
- `timeout`: The maximum timeout time when downloading data.

When loading the example data, the single-cell RNA **Seurat object** (preprocessed), the bulk RNA expression matrix, and the clinical phenotype data are loaded all at once. These data are combined into a list and returned.

```{r}
#| label: load_ref_data_example
#| eval: false
mydata <- LoadRefData(
  data_type = c("survival"),
  path = tempdir(),
  cache = TRUE,
  timeout = 60
)

# * seurat_obj (single-cell data, preprocessed)
mydata[[1]]
# An object of class Seurat
# 15924 features across 1077 samples within 1 assay
# Active assay: RNA (15924 features, 2000 variable features)
#  3 layers present: counts, data, scale.data
#  2 dimensional reductions calculated: pca, umap

# * bulk_survival
mydata[[2]][1:6, 1:6]
#         TCGA-69-7978 TCGA-62-8399 TCGA-78-7539 TCGA-73-4658 TCGA-44-6775 TCGA-44-2655
# HIF3A         4.2598      11.6239       9.1362       5.0288       4.0573       5.5335
# RTN4RL2       8.2023       5.5819       3.5365       7.4156       7.7107       5.3257
# HMGCLL1       2.7476       5.8513       3.8334       3.6447       2.9188       4.8820
# LRRTM1        0.0000       0.4628       4.7506       6.8005       7.7819       2.2882
# GRIN1         6.6074       5.4257       4.9563       7.3510       3.5361       3.3311
# LRRTM3        1.7458       2.0092       0.0000       1.4468       0.0000       0.0000

# * pheno_survival
mydata[[3]] |> head()
#               time status
# TCGA-69-7978  4.40      0
# TCGA-62-8399 88.57      0
# TCGA-78-7539 25.99      0
# TCGA-73-4658 52.56      1
# TCGA-44-6775 23.16      0
# TCGA-44-2655 43.50      0
```

We recommend using the `zeallot` package's `%<-%` function to assign values and rename them simultaneously.

```{r}
#| label: load_ref_data_example_2
#| eval: false
library(zeallot)

c(seurat_obj, bulk, pheno) %<-%
  LoadRefData(
    data_type = c("survival"),
    path = tempdir(),
    cache = TRUE,
    timeout = 60
  )
```

## Aggregating Duplicate Rows or Columns

These functions collapse duplicated row names (e.g., gene symbols) or column names (e.g., sample IDs) in expression matrices and count tables using configurable aggregation methods.

### Core Functions
| Function | Purpose |
|----------|---------|
| `AggregateDupRows()` | Merge rows with identical names (e.g., duplicate genes) |
| `AggregateDupCols()` | Merge columns with identical names (e.g., duplicate samples) |
| `AggregateDups()` | Convenience wrapper: merge rows & columns |

### Supported Methods
- `"max"` (default), `"sum"`, `"mean"`, `"median"`, `"first"`

### Basic Usage

```{r}
#| eval: false
# Example matrix with duplicate genes (rows) and samples (columns)
mat <- matrix(
  1:16,
  nrow = 4,
  dimnames = list(c("TP53", "TP53", "BRCA1", "ACTB"), c("S1", "S1", "S2", "S3"))
)

mat
#       S1 S1 S2 S3
# TP53   1  5  9 13
# TP53   2  6 10 14
# BRCA1  3  7 11 15
# ACTB   4  8 12 16

# Collapse duplicate genes using sum
AggregateDupRows(mat, method = "sum")
#       S1 S1 S2 S3
# TP53   2  6 10 14
# BRCA1  3  7 11 15
# ACTB   4  8 12 16

# Collapse duplicate samples using max
AggregateDupCols(mat, method = "max")
#       S1 S2 S3
# TP53   5  9 13
# TP53   6 10 14
# BRCA1  7 11 15
# ACTB   8 12 16

# Full deduplication in one step
AggregateDups(mat, method = "sum")
#       S1 S2 S3
# TP53  14 19 27
# BRCA1 10 11 15
# ACTB  12 12 16
```

### Features

- ✅ Handles `matrix`, `data.frame`, and S4 `Matrix` classes (e.g., `dgCMatrix`)
- ✅ Preserves original order of *first occurrence* for each unique name
- ✅ Silent mode: `verbose = FALSE`
- ✅ Independent row/column methods via `AggregateDups(row_method, col_method)`

## Standalone

### get_var_value

`get_var_value` is a **standalone script** that statically traces
variable assignments inside an R function (both from function parameters
and the function body) and computes the variable's value by recursively
resolving the dependency chain.

#### Example

```{r}
#| label: get_var_value_example
#| eval: false
g <- function(a = 1, b = 2) {
  c <<- a * 2 + b * 3
  d <- c^2
  e <- d - 1
  e
}
get_var_value("d", g) # returns 64

h <- function(a = "A", ...) {
  a <- 1
  return(a)
  a <- 2
  a
}
get_var_value("a", h) # returns 1
```

To copy `get_var_value` into your own package, run:

```{r}
#| label: usethis_use_standalone_example
#| eval: false
usethis::use_standalone("WangLabCSU/SigBridgeR", "standalone-get_var_value.R")
```

### get_env_vars

`get_env_vars` is a **standalone script** that captures every binding
(including hidden variables whose names start with `.`) from a given
environment and returns them as a **named list**. It is self-contained
(only uses `base` functions) and can be sourced independently or imported
via `source()` / `box::use()` without pulling in additional package
dependencies.

Internally, it calls `ls(envir = env, all.names = all.names)` to list the
variable names, then `mget(vars, envir = env, inherits = FALSE)` to
retrieve the corresponding values. `inherits = FALSE` ensures that only
bindings directly defined in `env` are returned, not those inherited from
enclosing frames. By default, `env = parent.frame()`, which captures the
caller's environment — the most common use case inside a function body.

Typical use cases:

- **Debugging**: snapshot the current local environment state.
- **Metaprogramming**: capture all objects created inside a function's body.
- **Caching / reproducibility**: serialise the environment to disk for later
  inspection or replay.

#### Parameters

- `env`: An environment whose bindings should be captured. Defaults to
  `parent.frame()`, i.e. the calling environment.
- `all.names`: Logical. If `TRUE`, all object names are returned; if
  `FALSE`, names which begin with a `.` are omitted. Default is `FALSE`.
- `exclude`: A character vector of variable names to exclude from the
  result. Default is `character(0)` (no exclusion).

#### Example

```{r}
#| label: get_env_vars_example
#| eval: false
source("R/standalone-get_env_vars.R")

f <- function() {
  x <- 1
  y <- "hello"
  .hidden <- "not shown by default"
  get_env_vars()
}
f()
# $x
# [1] 1
#
# $y
# [1] "hello"

# Also capture hidden variables starting with "."
g <- function() {
  x <- 1
  .hidden <- "shown"
  get_env_vars(all.names = TRUE)
}
g()
# $x
# [1] 1
#
# $.hidden
# [1] "shown"
```

To copy `get_env_vars` into your own package, run:

```{r}
#| label: usethis_use_standalone_get_env_vars
#| eval: false
usethis::use_standalone("WangLabCSU/SigBridgeR", "standalone-get_env_vars.R")
```

