---
title: "Get Started with SigBridgeR"
date: '`r Sys.Date()`'
output: 
  html_document
knit: knitted
vignette: >
  %\VignetteIndexEntry{Get Started with SigBridgeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Preparation

```{r}
#| label: prework
#| include: false
#| eval: false

library(SigBridgeR)
library(Seurat)
library(zeallot)
```

We will start with a simple example.

```{r}
#| label: example_data
#| eval: false
ref <- LoadRefData(data_type = "binary")
seurat_obj <- ref[[1]]
bulk_bi <- ref[[2]]
pheno_bi <- ref[[3]]
```

`seurat_obj` is a preprocessed single-cell RNA **Seurat** object (QC filtering, normalization, clustering, and UMAP are already included), `bulk_bi` is a bulk tissue RNA expression matrix (TPM), and `pheno_bi` is the phenotypic data associated with `bulk_bi`. When using a binary or continuous phenotype, the reference phenotype data is a **named vector**.

```{r}
#| label: binary_phenotype
#| eval: false
head(pheno_bi)
```

By the way, when using patients' survival data as phenotype, the reference data is a **data.frame**.

```{r}
#| label: survival_phenotype
#| eval: false
pheno_sur <- LoadRefData(data_type = "survival")[[3]]
head(pheno_sur)
```

The reference single-cell RNA data has already been preprocessed into a **Seurat** object (`seurat_obj`), so it can be passed directly to `Screen()` without any further processing.

## Screening

Then we can use these data to screen out phenotype-assoicated cells. Let's start by trying **Scissor**.

```{r}
#| label: scissor
#| eval: false
scissor_res <- Screen(
  bulk_bi,
  seurat_obj,
  pheno_bi,
  phenotype_class = "binary",
  screen_method = "Scissor",
  alpha = 0.05
)
```

```{r}
#| eval: false
scissor_res |> class()
```

```{r}
#| eval: false
scissor_res |> print()
```

Other screening methods are also available.

```{r}
#| label: other_screening
#| eval: false
scpas_res <- Screen(
  bulk_bi,
  seurat_obj,
  pheno_bi,
  phenotype_class = "binary",
  screen_method = "scPAS",
  alpha = 0.05
)
```

Since the screening is performed on the same data, we can merge them based on the cell names. Metadata will be merged as well.

```{r}
#| label: merge
#| eval: false
merged_seurat <- MergeResult(
  scissor_res,
  scpas_res
)
```

## Visualization

Finally, we can visualize the screening results.

-   stacked bar plot:

```{r}
#| label: visualize_fraction
#| eval: false
fraction <- ScreenFractionPlot(
  merged_seurat,
  group_by = "seurat_clusters",
  screen_type = c("scissor", "scPAS")
)

names(fraction)

fraction$combined_plot
```


-   Venn diagram:

```{r}
#| label: visualize_venn
#| eval: false
c(scissor_pos, scpas_pos) %<-%
  purrr::map(
    c("scissor", "scPAS"),
    ~ colnames(merged_seurat)[
      which(merged_seurat[[.x]] == "Positive")
    ]
  )

all_cells <- colnames(seurat_obj)

pos_venn <- list(
  scissor = scissor_pos,
  scpas = scpas_pos,
  all_cells = all_cells
)

set.seed(123)

venn_plot <- ggVennDiagram::ggVennDiagram(
  x = pos_venn,
  # * the labels of each group to be shown on the diagram
  category.names = c(
    "Scissor",
    "scPAS",
    "All cells"
  ),
  # * the colors of each group
  set_color = c(
    "#a33333ff",
    "#37ae00ff",
    "#008383ff"
  ),
  label_geom = c("text")
) +
  ggplot2::scale_fill_gradient(low = "white", high = "#ffb6b6ff") +
  ggplot2::ggtitle("Screening Venn Diagram")

venn_plot
```


-   Set plot:

```{r}
#| label: visualize_set
#| eval: false
upset <- ScreenUpset(
  merged_seurat,
  screen_type = c("scissor", "scPAS")
)
names(upset)

upset$plot
```

-   2D UMAP:

```{r}
#| label: visualize_umap
#| eval: false
library(patchwork)
library(randomcoloR)

c(
  scissor_umap,
  scpas_umap
) %<-%
  purrr::map(
    c("scissor", "scPAS"),
    ~ Seurat::DimPlot(
      merged_seurat,
      group.by = .x,
      pt.size = 0.1,
      reduction = "umap",
      cols = c(
        "Neutral" = "#CECECE",
        "Positive" = "#ff3333",
        "Negative" = "#386c9b"
      )
    ) +
      ggplot2::ggtitle(.x)
  )

set.seed(123)
cols <- randomcoloR::distinctColorPalette(
  length(unique(merged_seurat$seurat_clusters)),
  runTsne = TRUE
)

cluster_umap <- Seurat::DimPlot(
  merged_seurat,
  group.by = "seurat_clusters",
  pt.size = 0.1,
  reduction = "umap",
  cols = cols
) +
  ggplot2::ggtitle("seurat_clusters")


# * Show
umaps <- cluster_umap +
  scissor_umap +
  scpas_umap +
  plot_layout(ncol = 2)

umaps
```

UMAP has been registered in the `autoplot()` for easier check.

```{r}
#| label: umaps_q2
#| eval: false
ggplot2::autoplot(merged_seurat, group.by = "scPAS")
```

For more detail, please see [Full Tutorial](https://wanglabcsu.github.io/SigBridgeR/articles/Full_Tutorial.html).

## Session information:

```{r}
#| label: session_info
#| include: false
sessionInfo()
```
