---
title: "Troubleshooting"
author: "Exceret"
date: '`r Sys.Date()`'
output: 
  html_document: 
    toc: false  
knit: knitted
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
sidebar: false
---

# Overview

This vignette provides solutions to common problems you might encounter when using this package. If you cannot find your problem here, please file an issue on our GitHub repository.

## Scissor

> **Error in normalize.quantiles(dataset0) : ERROR; return code from pthread_create() is 22**

To solve this problem, you need to install `preprocessCore` without threading support. Try:

```shell
# ! shell
git clone https://github.com/bmbolstad/preprocessCore.git
cd preprocessCore
R CMD INSTALL --configure-args="--disable-threading"  .
```

or

```{r}
#| eval: false
BiocManager::install(
  "preprocessCore",
  configure.args = "--disable-threading",
  force = TRUE
)
```

See [bioconductor_docker/issues/22](https://github.com/Bioconductor/bioconductor_docker/issues/22), [Scissor/issues/15](https://github.com/sunduanchen/Scissor/issues/15) for more details.


> Error at alpha=0.05:subscript out of bounds
>
> Error in `Scissor.v5.optimized()`:
>
> ! object 'fit0' not found

This may be due to two reasons: first, a mismatch in the dimensions of the bulk expression data and the phenotype data; second, incorrect column names in the survival phenotype data leading to a failure to match.

To check dimension mismatch：

```{r}
#| eval: false
ncol(your_bulk_data) == nrow(your_phenotype_data) # should be TRUE

all(unique(colnames(your_bulk_data)) == unique(rownames(your_bulk_data))) # should be TRUE

all(order(colnames(your_bulk_data)) == order(rownames(your_phenotype_data))) # should be TRUE
```

Survival phenotype column names should be formatted as `time` and `status`, ensuring correct capitalization and spelling:

```{r}
#| eval: false
head(survival_phenotype) # case-sensitive
#           time status
# GSM70130 34.80      0
# GSM70131 35.67      0
# GSM70136 43.37      0
# GSM70138 60.77      0
# GSM70140 33.80      1
# GSM70144 58.53      0
```


## Package

> Error in `function_name()`:
>
> ! lazy-load database '/home/user/R/x86_64-pc-linux-gnu-library/4.4/SigBridgeR/R/SigBridgeR. rdb' is corrupt

An error occurred during installation, causing the package to be corrupted. Try reinstalling the package:

```{r}
#| eval: false
# I reccomend restarting R/RStudio before reinstalling
remove.packages("SigBridgeR")

detach("package:SigBridgeR", unload = TRUE)

if (!requireNamespace("remotes")) {
  install.packages("remotes")
}
remotes::install_github("WangLabCSU/SigBridgeR")
```

<br>


## Visualization

> Error:
>
> ! Invalid syntax: 'c(scissor_umap, scpas_umap)'

This may be due to the R environment being contaminated, which prevents the use of `%<-%`. You can try restarting the R environment or clean up the environment.

```{r}
#| label: restart_R
#| eval: false
# restart R/RStudio
.rs.api.restartSession()
```

```{r}
#| label: clean_all
#| eval: false
rm(list = ls(all.names = TRUE))
```

<br>


## scPP

> Error:
>
> ! Detected n gene(s) with zero variance:
>
> ℹ "gene name(s)"

This is due to the presence of genes with zero variance in the bulk expression data when you are using `scPP` and `binary` phenotype. This indicates that the expression levels of one (or several) genes are nearly identical across different samples. You should check your data.


>  ✖ $$2025/09/23 08:53:51$$ Fewer than 20% of the genes in the gene sets are included in the rankings.Check wether the gene IDs in the 'rankings' and 'geneSets' match.
>
> ** **ℹ $$2025/09/23 08:53:51$$ scPP screening exit.****

This issue arises from the single-cell processing, which filtered out too many genes and cells. Consider Adjusting `min.cells` and `min.features` to a smaller value.:




## Seurat

> Excessive time during running SCTransfomm

Output message is probably like this:

```{r}
#| eval: false
# Running SCTransform on assay: Xenium
# vst.flavor='v2' set. Using model with fixed slope and excluding poisson genes.
# Calculating cell attributes from input UMI matrix: log_umi
# Variance stabilizing transformation of count matrix of size 248 by 36553
# Model formula is y ~ log_umi
# Get Negative Binomial regression parameters per gene
# Using 248 genes, 5000 cells
# Second step: Get residuals using fitted parameters for 248 genes
# Computing corrected count matrix for 248 genes
# Calculating gene attributes
# Wall clock passed: Time difference of 7.166213 secs
# Determine variable features
# ! stuck here.
```

This is a bug from Seurat, see [Seurat #10153](https://github.com/satijalab/seurat/issues/10153)

Try update `Seurat` to latest version
